---
title: "ggplot2 Basics"
subtitle: "Data Visualization"
author: "YOUR NAME"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

# Getting Started

This document introduces the **ggplot2** workflow for building clear and professional visualizations in R.

You will learn how to:

- create scatterplots, line plots, and bar charts
- map variables to aesthetics (`aes()`)
- control color, size, transparency, and labels
- use `facet_wrap()` to compare groups
- polish plots with themes and annotations

<br>

# Setup

## Install (one-time)

```{r}
# install.packages("tidyverse")
```

## Load packages

```{r}
library(tidyverse)
```

<br>

# The ggplot2 Grammar of Graphics

Most ggplot charts follow this structure:

```{r}
#| eval: false
ggplot(data = DATA, aes(x = X_VAR, y = Y_VAR)) +
  geom_...( ) +
  labs(title = "...", x = "...", y = "...") +
  theme_minimal()
```

Key pieces:

- **Data**: a data frame
- **Aesthetics (`aes`)**: how variables map to visual elements
- **Geoms**: the type of plot (points, lines, bars, etc.)
- **Labels**: title, subtitle, axes, caption
- **Theme**: overall style

<br>

# Example Dataset

We will use the built-in dataset `mpg` (fuel economy data).

```{r}
mpg |> glimpse()
```

<br>

# Scatterplots

## Basic scatterplot

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Add transparency

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5)
```

## Map color to a categorical variable

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7)
```

## Map size to a numeric variable

```{r}
ggplot(mpg, aes(x = displ, y = hwy, size = cyl)) +
  geom_point(alpha = 0.6)
```

✅ Rule of thumb:  
- `color = ...` inside `aes()` means **it changes by data values**  
- `color = "blue"` outside `aes()` means **fixed color**

<br>

# Trend Lines

## Add a smooth trend

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  geom_smooth()
```

## Linear regression line

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE)
```

<br>

# Faceting (Small Multiples)

Facets create a grid of plots split by a group variable.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ class)
```

<br>

# Bar Charts

Bar charts are usually used for **categorical variables**.

## Count bar chart

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
```

## Flip coordinates (nice for long labels)

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar() +
  coord_flip()
```

## Bar chart with fill color

```{r}
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar()
```

<br>

# Summarized Bar Charts

Sometimes you want bars representing *summary statistics* (means, totals, etc.).

## Step 1: Summarize with dplyr

```{r}
mpg_mean <- mpg |>
  group_by(class) |>
  summarise(mean_hwy = mean(hwy), .groups = "drop")

mpg_mean
```

## Step 2: Plot the summarized data

```{r}
ggplot(mpg_mean, aes(x = class, y = mean_hwy)) +
  geom_col()
```

✅ `geom_bar()` counts rows automatically.  
✅ `geom_col()` uses your own y-values.

<br>

# Line Charts

Line charts require an x-variable that has a meaningful order (often time).

We will build a small example dataset.

```{r}
sales <- tibble(
  month = 1:12,
  revenue = c(10, 12, 11, 13, 15, 18, 17, 20, 22, 23, 25, 28)
)

sales
```

## Basic line plot

```{r}
ggplot(sales, aes(x = month, y = revenue)) +
  geom_line()
```

## Line + points

```{r}
ggplot(sales, aes(x = month, y = revenue)) +
  geom_line() +
  geom_point()
```

<br>

# Labels and Titles

Use `labs()` to make your charts understandable.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Engine Size vs Highway MPG",
    subtitle = "Bigger engines usually get lower fuel efficiency",
    x = "Engine displacement (liters)",
    y = "Highway MPG",
    caption = "Source: ggplot2::mpg"
  )
```

<br>

# Themes

Themes change the overall style.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  theme_minimal()
```

Try a few common ones:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  theme_classic()
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6) +
  theme_light()
```

<br>

# Changing Colors (Manually)

If you want a **custom palette**, you can set:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7) +
  scale_color_brewer(palette = "Set2")
```

Or use a fixed color:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.7, color = "steelblue")
```

<br>

# Saving Plots

Use `ggsave()` to export a plot.

```{r}
p <- ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.6)

ggsave("my_scatterplot.png", plot = p, width = 7, height = 5, dpi = 300)
```

<br>

# Practice Problems ✅

1. Create a scatterplot of `cty` vs `hwy`.  
2. Color points by `drv` and set `alpha = 0.6`.  
3. Add a linear trend line (`method = "lm"`, `se = FALSE`).  
4. Use `facet_wrap(~ drv)` to compare groups.  
5. Create a bar chart of `manufacturer`. Flip the axis using `coord_flip()`.  
6. Create a summarized bar chart of mean `hwy` by `manufacturer` (**top 10 only**).  

<br>

# Challenge 💡 (Top 10 Manufacturers)

```{r}
top10 <- mpg |>
  count(manufacturer, sort = TRUE) |>
  slice_head(n = 10)

top10
```

Now:

- filter `mpg` to only these 10 manufacturers
- compute mean `hwy` by manufacturer
- make a `geom_col()` plot
- flip the coordinates and add titles
