---
title: "dplyr Basics"
subtitle: "Data Transformation"
author: "YOUR NAME"
date: last-modified
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show the code"
    highlight-style: atom-one
execute:
  echo: true
  warning: false
  message: false
---

# Getting Started

This document introduces the **dplyr** workflow for cleaning, transforming, and summarizing data in R.

You will learn how to:

- select columns with `select()`
- filter rows with `filter()`
- sort data with `arrange()`
- create new variables with `mutate()`
- summarize data with `summarise()`
- group operations with `group_by()`
- combine datasets with joins

<br>

# Setup

## Install (one-time)

```{r}
# install.packages("tidyverse")
```

## Load packages

```{r}
library(tidyverse)
```

<br>

# A Quick Look at a Dataset

We will use the built-in dataset **`mpg`**.

```{r}
mpg |> glimpse()
```

<br>

# The Pipe Operator `|>`

The pipe sends the output of one step into the next step.

```{r}
mpg |>
  select(manufacturer, model, year)
```

This is equivalent to:

```{r}
select(mpg, manufacturer, model, year)
```

✅ Use pipes to write code in a clear, step-by-step style.

<br>

# `select()` (Choose Columns)

## Select a few columns

```{r}
mpg |>
  select(manufacturer, model, displ, hwy)
```

## Select a range of columns

```{r}
mpg |>
  select(manufacturer:year)
```

## Remove columns

```{r}
mpg |>
  select(-cty, -hwy)
```

<br>

# `filter()` (Choose Rows)

## Filter by a condition

```{r}
mpg |>
  filter(class == "compact")
```

## Multiple conditions with AND

```{r}
mpg |>
  filter(class == "compact", year == 2008)
```

## OR conditions using `%in%`

```{r}
mpg |>
  filter(class %in% c("compact", "suv"))
```

<br>

# `arrange()` (Sort Rows)

## Sort ascending

```{r}
mpg |>
  arrange(hwy)
```

## Sort descending

```{r}
mpg |>
  arrange(desc(hwy))
```

## Sort by multiple variables

```{r}
mpg |>
  arrange(class, desc(hwy))
```

<br>

# `mutate()` (Create New Variables)

## Create a new variable

```{r}
mpg |>
  mutate(hwy_per_liter = hwy / displ)
```

## Create multiple new variables

```{r}
mpg |>
  mutate(
    hwy_per_liter = hwy / displ,
    cty_per_liter = cty / displ
  )
```

## Use `if_else()` for conditional logic

```{r}
mpg |>
  mutate(
    big_engine = if_else(displ >= 4, "Yes", "No")
  ) |>
  count(big_engine)
```

<br>

# `summarise()` (Compute Summary Statistics)

## One summary value

```{r}
mpg |>
  summarise(mean_hwy = mean(hwy))
```

## Multiple summaries

```{r}
mpg |>
  summarise(
    mean_hwy = mean(hwy),
    sd_hwy   = sd(hwy),
    max_hwy  = max(hwy),
    min_hwy  = min(hwy)
  )
```

✅ Tip: Use `na.rm = TRUE` if your data has missing values.

<br>

# `group_by()` + `summarise()` (Grouped Summaries)

## Mean highway MPG by class

```{r}
mpg |>
  group_by(class) |>
  summarise(mean_hwy = mean(hwy), .groups = "drop") |>
  arrange(desc(mean_hwy))
```

## Count rows by group

```{r}
mpg |>
  group_by(drv) |>
  summarise(n = n(), .groups = "drop")
```

<br>

# `count()` (Fast Frequency Tables)

```{r}
mpg |>
  count(class, sort = TRUE)
```

With two variables:

```{r}
mpg |>
  count(class, drv, sort = TRUE)
```

<br>

# `distinct()` (Unique Values)

```{r}
mpg |>
  distinct(manufacturer)
```

Unique combinations:

```{r}
mpg |>
  distinct(manufacturer, year)
```

<br>

# `slice_*()` (Pick Specific Rows)

## Top rows

```{r}
mpg |>
  slice_head(n = 5)
```

## Bottom rows

```{r}
mpg |>
  slice_tail(n = 5)
```

## Top rows by a variable

```{r}
mpg |>
  arrange(desc(hwy)) |>
  slice_head(n = 10)
```

<br>

# `across()` (Apply Functions to Many Columns)

Example: compute means for multiple numeric columns.

```{r}
mpg |>
  summarise(
    across(c(cty, hwy, displ), mean)
  )
```

You can combine with `group_by()`:

```{r}
mpg |>
  group_by(class) |>
  summarise(
    across(c(cty, hwy), mean),
    .groups = "drop"
  )
```

<br>

# Joining Data (Combining Tables)

Joins combine tables using a matching key column.

We will create two small example tables.

```{r}
students <- tibble(
  id = c(101, 102, 103),
  name = c("Alex", "Bella", "Chris")
)

grades <- tibble(
  id = c(101, 103),
  score = c(95, 88)
)
```

## `left_join()` (most common)

```{r}
students |>
  left_join(grades, by = "id")
```

## `inner_join()`

```{r}
students |>
  inner_join(grades, by = "id")
```

## `anti_join()` (find non-matches)

```{r}
students |>
  anti_join(grades, by = "id")
```

<br>

# A Typical dplyr Workflow Example

Goal: find the best average highway MPG by car class (with sample size).

```{r}
mpg |>
  group_by(class) |>
  summarise(
    mean_hwy = mean(hwy),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(mean_hwy))
```

<br>

# Practice Problems ✅

1. Use `select()` to keep only `manufacturer`, `model`, `year`, `hwy`.  
2. Filter to only cars with `drv == "f"` and `year == 2008`.  
3. Sort the filtered data by `hwy` (descending).  
4. Create a new variable `efficiency = hwy / displ`.  
5. Compute the mean of `efficiency`.  
6. Compute mean `hwy` by `class` and sort from highest to lowest.  
7. Find the top 5 classes by mean `hwy`.  

<br>

# Challenge 💡 (Top Manufacturers)

1. Count how many rows each manufacturer has.
2. Keep only manufacturers with **at least 20 cars** in the dataset.
3. Compute mean highway MPG by manufacturer.
4. Sort the results from highest to lowest.

Starter code:

```{r}
mpg |>
  count(manufacturer, sort = TRUE)
```
